---
title: "Local Analysis of Plant Genomes with PlantTxDbHub"
author: "Kabilan S"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Local Analysis of Plant Genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PlantTxDbHub)
library(GenomicFeatures)
library(GenomeInfoDb)

# Only download the species actually used later in this vignette, and do
# so defensively: a vignette must not hard-fail if the network is
# unavailable during package build/check (see Bioconductor "Querying Web
# Resources" guideline).
species_used <- c("Arabidopsis_TAIR10", "Glycine_Gmax_v2.1", "Oryza_IRGSP_1")

db_dir <- tryCatch(
  downloadPlantTxDbs(species = species_used),
  error = function(e) {
    message("Could not download TxDb files: ", conditionMessage(e))
    NA
  }
)

data_available <- !is.na(db_dir) &&
  all(vapply(species_used, function(sp) {
    file.exists(tryCatch(getTxDbPath(sp, dest_dir = db_dir),
      error = function(e) NA_character_
    ))
  }, logical(1)))
```

# Installation

## Install from Bioconductor

Once **PlantTxDbHub** is accepted into Bioconductor, install it with:

```{r install_bioconductor, eval = FALSE}
if (!require("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("PlantTxDbHub")
```

## Install the development version from GitHub

While the package is under Bioconductor review, install the development
version directly from GitHub:

```{r install_github, eval = FALSE}
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("kabilanbio/PlantTxDbHub")
```

# Introduction

**PlantTxDbHub** provides ready‑to‑use TxDb (transcript database) annotations
for plant genomes. The databases are stored as SQLite files and can be
downloaded on demand using `downloadPlantTxDbs()`.
The available species and their download URLs are maintained in a curated
CSV file (`inst/extdata/metadata.csv`). This design allows the community to
add new species without modifying any R code.

# Finding available species

Use `listPlantTxDbSpecies()` to see the available species and their
associated metadata (species name, genome build, source version, data
provider, filename, and description):

```{r list_species}
listPlantTxDbSpecies()
```

# Downloading the databases

`downloadPlantTxDbs()` reads the metadata CSV and downloads each SQLite file
from its unique URL, caching it via `BiocFileCache`. By default it fetches
**all** available databases; you can also request a subset by `SpeciesID`.

**Download all databases:**

```{r download_all, eval = FALSE}
downloadPlantTxDbs()
```

**Download only specific species** (as done in the setup chunk above):

```{r download_single, eval = FALSE}
# Only Arabidopsis
downloadPlantTxDbs(species = "Arabidopsis_TAIR10")

# Arabidopsis and rice together
downloadPlantTxDbs(species = c("Arabidopsis_TAIR10", "Oryza_IRGSP_1"))
```

Files are cached in a persistent user data directory via `BiocFileCache`, so
subsequent calls are fast and corrupted files are re-downloaded
automatically.

# Loading a TxDb (Arabidopsis example)

Once downloaded, obtain the full path to a cached file with
`getTxDbPath()`, then load it with `GenomicFeatures::loadDb()`.

The remaining sections of this vignette require a successful download; they
are skipped if the network was unavailable during vignette build.

```{r load_ath, eval = data_available}
ath_file <- getTxDbPath("Arabidopsis_TAIR10")
txdb_ath <- loadDb(ath_file)
txdb_ath
```

# Available columns and keys

TxDb objects support the `select()` interface.

```{r cols, eval = data_available}
columns(txdb_ath)
keytypes(txdb_ath)
```

The column **`TXTYPE`** indicates the transcript biotype (e.g. `protein_coding`).

# Retrieve all genes

`genes()` returns a `GRanges` with gene‑level information.

```{r genes_ath, eval = data_available}
gene_gr <- genes(txdb_ath)
head(gene_gr)
```

# Retrieve all transcripts

```{r transcripts_ath, eval = data_available}
tx_gr <- transcripts(txdb_ath)
head(tx_gr)
```

# Retrieve all exons

```{r exons_ath, eval = data_available}
ex_gr <- exons(txdb_ath, columns = "exon_id")
head(ex_gr)
```

# Filter by gene ID

```{r filter_geneid, eval = data_available}
my_genes <- c("AT1G01010", "AT1G01020")
genes(txdb_ath, filter = list(gene_id = my_genes))
```

# Retrieving transcript types

```{r txtype, eval = data_available}
tx_info <- select(txdb_ath,
  keys = keys(txdb_ath, "TXID"),
  columns = c("TXID", "TXTYPE"),
  keytype = "TXID"
)
table(tx_info$TXTYPE)
```

# Working with chromosome names

```{r chrom_ath, eval = data_available}
gene_gr_nuc <- keepSeqlevels(gene_gr,
  value = c("1", "2", "3", "4", "5"),
  pruning.mode = "coarse"
)
seqlevels(gene_gr_nuc) <- paste0("Chr", seqlevels(gene_gr_nuc))
seqlevels(gene_gr_nuc)
```

# Soybean (*Glycine max*) example

Two soybean genome builds are available: `Glycine_Gmax_v2.1` (Ensembl
Plants) and `Glycine_Wm82.a4.v1` (Phytozome). This example uses the Ensembl
Plants build.

```{r load_gmx, eval = data_available}
gmx_file <- getTxDbPath("Glycine_Gmax_v2.1")
txdb_gmx <- loadDb(gmx_file)
head(genes(txdb_gmx))
```

# Rice (*Oryza sativa*) example

```{r load_osa, eval = data_available}
osa_file <- getTxDbPath("Oryza_IRGSP_1")
txdb_osa <- loadDb(osa_file)
head(genes(txdb_osa))
```

# Contributing new species

We welcome the addition of TxDb databases for other plant species.
To contribute:

1. **Host the SQLite file** on a permanent, publicly accessible URL
   (e.g., your own Zenodo record).
2. **Fork** the package repository and edit the file
   `inst/extdata/metadata.csv`.
3. **Add a new row** containing:
   - `SpeciesID` – a unique identifier (e.g. `Zea_mays_B73`).
   - `Title` – the exact filename of the SQLite file.
   - `Location_Prefix` and `RDataPath` – the two parts of the download URL
     (they will be concatenated to form the full link).
   - Other columns as appropriate (`Description`, `Genome`, `Species`,
     `DataProvider`, etc.).
   Follow the existing rows for guidance.
4. **Submit a pull request** with your changes. Once merged and a new
   version of the package is released, the new species will be instantly
   available to all users.

No R code changes are required – the package reads all necessary
information directly from `metadata.csv`.

# Data sources and licensing

TxDb resources in this package are derived from Ensembl Plants and
Phytozome annotation data; see `?PlantTxDbHub` for full licensing details
and `getTxDbProvenance()` to query provenance per species.

# Session information

```{r sessionInfo}
sessionInfo()
```
