---
title: "Local Analysis of Plant Genomes with PlantTxDbHub"
author: "Kabilan S"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Local Analysis of Plant Genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(PlantTxDbHub)
library(GenomicFeatures)
library(GenomeInfoDb)

# Download all available databases
db_dir <- downloadPlantTxDbs()
```

# Installation

Install the development version from GitHub:

```{r install, eval = FALSE}
if (!require("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("kabilanbio/PlantTxDbHub")
```

# Introduction

**PlantTxDbHub** provides ready‑to‑use TxDb (transcript database) annotations
for plant genomes. The databases are stored as SQLite files and can be
downloaded on demand using `downloadPlantTxDbs()`.  
The available species and their download URLs are maintained in a curated
CSV file (`inst/extdata/metadata.csv`). This design allows the community to
add new species without modifying any R code.

# Finding available species

Use `listPlantTxDbSpecies()` to see the species identifiers and their
corresponding SQLite file names:

```{r list_species}
listPlantTxDbSpecies()
```

# Downloading the databases

`downloadPlantTxDbs()` reads the metadata CSV and downloads each SQLite file
from its unique URL. By default it fetches **all** databases; you can also
request a subset.

**Download all databases** (already done in the setup chunk):

```{r download_all, eval = FALSE}
downloadPlantTxDbs()
```

**Download only specific species:**

```{r download_single, eval = FALSE}
# Only Arabidopsis
downloadPlantTxDbs(species = "Arabidopsis_TAIR10")

# Arabidopsis and rice together
downloadPlantTxDbs(species = c("Arabidopsis_TAIR10", "Oryza_IRGSP_1"))
```

The files are cached in a persistent user data directory, so subsequent calls
are instantaneous unless a file is corrupted.

# Loading a TxDb (Arabidopsis example)

Once downloaded, you can obtain the full path to a cached file with
`getTxDbPath()`, then load it with `GenomicFeatures::loadDb()`:

```{r load_ath}
ath_file <- getTxDbPath("Arabidopsis_TAIR10")
txdb_ath <- loadDb(ath_file)
txdb_ath
```

# Available columns and keys

TxDb objects support the `select()` interface.

```{r cols}
columns(txdb_ath)
keytypes(txdb_ath)
```

The column **`TXTYPE`** indicates the transcript biotype (e.g. `protein_coding`).

# Retrieve all genes

`genes()` returns a `GRanges` with gene‑level information.

```{r genes_ath}
gene_gr <- genes(txdb_ath)
head(gene_gr)
```

# Retrieve all transcripts

```{r transcripts_ath}
tx_gr <- transcripts(txdb_ath)
head(tx_gr)
```

# Retrieve all exons

```{r exons_ath}
ex_gr <- exons(txdb_ath, columns = "exon_id")
head(ex_gr)
```

# Filter by gene ID

```{r filter_geneid}
my_genes <- c("AT1G01010", "AT1G01020")
genes(txdb_ath, filter = list(gene_id = my_genes))
```

# Retrieving transcript types

```{r txtype}
tx_info <- select(txdb_ath,
  keys = keys(txdb_ath, "TXID"),
  columns = c("TXID", "TXTYPE"),
  keytype = "TXID"
)
table(tx_info$TXTYPE)
```

# Working with chromosome names

```{r chrom_ath}
gene_gr_nuc <- keepSeqlevels(gene_gr,
  value = c("1", "2", "3", "4", "5"),
  pruning.mode = "coarse"
)
seqlevels(gene_gr_nuc) <- paste0("Chr", seqlevels(gene_gr_nuc))
seqlevels(gene_gr_nuc)
```

# Soybean (*Glycine max*) example

```{r load_gmx}
gmx_file <- getTxDbPath("Glycine_Gmax_v2.1")
txdb_gmx <- loadDb(gmx_file)
head(genes(txdb_gmx))
```

# Rice (*Oryza sativa*) example

```{r load_osa}
osa_file <- getTxDbPath("Oryza_IRGSP_1")
txdb_osa <- loadDb(osa_file)
head(genes(txdb_osa))
```

# Contributing new species

We welcome the addition of TxDb databases for other plant species.
To contribute:

1. **Host the SQLite file** on a permanent, publicly accessible URL
   (e.g., your own Zenodo record).
2. **Fork** the package repository and edit the file
   `inst/extdata/metadata.csv`.
3. **Add a new row** containing:
   - `SpeciesID` – a unique identifier (e.g. `Zea_mays_B73`).
   - `Title` – the exact filename of the SQLite file.
   - `Location_Prefix` and `RDataPath` – the two parts of the download URL
     (they will be concatenated to form the full link).
   - Other columns as appropriate (Description, Genome, Species, etc.).
   Follow the existing rows for guidance.
4. **Submit a pull request** with your changes. Once merged and a new
   version of the package is released, the new species will be instantly
   available to all users.

No R code changes are required – the package reads all necessary
information directly from `metadata.csv`.

# Session information

```{r sessionInfo}
sessionInfo()
```

