---
title: "Local Analysis of Plant Genomes with PlantTxDbHub"
author: "Kabilan S"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Local Analysis of Plant Genomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(PlantTxDbHub)
library(GenomicFeatures)
library(GenomeInfoDb)
```

# Introduction

This vignette demonstrates how to use the **PlantTxDbHub** package
to download and analyse transcript‑level annotations for three plant
species:

- *Arabidopsis thaliana* (TAIR10)
- *Glycine max* (Wm82 v2.1)
- *Oryza sativa* (IRGSP‑1.0)

The annotations were generated from Ensembl Plants release 62 GTF files
and stored as standalone TxDb SQLite databases on Zenodo.  
The package provides a helper function, `downloadPlantTxDbs()`, to obtain
these files locally; subsequent analysis uses standard `TxDb` methods.

# Download the TxDb databases

The three SQLite files are hosted on Zenodo. The following code creates
a local directory and downloads any missing databases. This step only
needs to be done once per machine.

```{r download, eval = FALSE}
library(PlantTxDbHub)
downloadPlantTxDbs() # downloads into "plant_txdb_sqlite/"
```

*Note*: In a future version the package will export `downloadPlantTxDbs()`
which wraps exactly this code.

# Load a TxDb (Arabidopsis example)

```{r load_ath}
# Point to the downloaded database
ath_file <- file.path("plant_txdb_sqlite", "TxDb.Athaliana.TAIR10.v62.sqlite")
if (!file.exists(ath_file)) {
  message("File not found. Please run the download chunk first.")
  knitr::knit_exit()
}
txdb_ath <- loadDb(ath_file)
txdb_ath
```

# Available columns and keys

TxDb objects work with the `select()` interface. You can discover the
available columns and key types:

```{r cols}
columns(txdb_ath)
keytypes(txdb_ath)
```

Common columns include `GENEID`, `TXID`, `EXONID`, `CDSID`, and
metadata columns like `GENEBIOTYPE`, `TXBIOTYPE`, `GENEID`, etc.
(availability depends on what was stored when the GTF was imported).

# Retrieve all genes

`genes()` returns a `GRanges` with gene‑level information. The metadata
columns depend on how the TxDb was constructed; for these GTF‑based
databases you will typically see `gene_id` and `gene_biotype`.

```{r genes_ath}
gene_gr <- genes(txdb_ath)
head(gene_gr)
```

# Retrieve all transcripts

```{r transcripts_ath}
tx_gr <- transcripts(txdb_ath, columns = c("tx_name", "gene_id", "tx_biotype"))
head(tx_gr)
```

# Retrieve all exons

```{r exons_ath}
ex_gr <- exons(txdb_ath, columns = "exon_id")
head(ex_gr)
```

# Filter by gene ID

Use `select()` to retrieve specific genes by their `GENEID`.  
Alternatively, subset the `GRanges` directly using the `gene_id` column.

```{r filter_geneid}
# Get the GRanges for a single gene
my_genes <- c("AT1G01010", "AT1G01020")
genes(txdb_ath, filter = list(gene_id = my_genes))
```

If the `filter` argument is not supported (older `GenomicFeatures`
versions), fall back to `select()`:

```{r filter_geneid_select}
# Retrieve gene ranges + metadata
sel <- select(txdb_ath,
              keys = my_genes,
              columns = c("GENEID", "TXID", "TXBIOTYPE", "EXONID"),
              keytype = "GENEID")
head(sel)
```

# Filter by gene biotype (e.g., protein‑coding)

```{r filter_biotype}
# Extract all gene IDs with gene_biotype "protein_coding"
gene_md <- select(txdb_ath,
                  keys = keys(txdb_ath, "GENEID"),
                  columns = c("GENEID", "GENEBIOTYPE"),
                  keytype = "GENEID")
pc_ids <- gene_md$GENEID[gene_md$GENEBIOTYPE == "protein_coding"]

# Subset the gene GRanges
pc_genes <- gene_gr[gene_gr$gene_id %in% pc_ids]
length(pc_genes)
head(pc_genes)
```

# Working with chromosome names

The database stores chromosome names as bare numbers (`1`, `2`, …)
and organelle names as `Mt`, `Pt`. To add the standard *Arabidopsis*
`Chr` prefix and restrict to nuclear chromosomes:

```{r chrom_ath}
gene_gr_nuc <- keepSeqlevels(gene_gr,
                             value = c("1", "2", "3", "4", "5"),
                             pruning.mode = "coarse")
seqlevels(gene_gr_nuc) <- paste0("Chr", seqlevels(gene_gr_nuc))
seqlevels(gene_gr_nuc)
```

# Soybean (*Glycine max*) example

The workflow is identical. The chromosome names in soybean are
`Gm01`, `Gm02`, …, `Gm20` (depending on the GTF). We show a quick
overview.

```{r load_gmx}
gmx_file <- file.path("plant_txdb_sqlite", "TxDb.Gmax.Wm82.v62.sqlite")
if (!file.exists(gmx_file)) {
  message("File not found. Run download chunk first.")
} else {
  txdb_gmx <- loadDb(gmx_file)
  head(genes(txdb_gmx))
  # List chromosome names
  seqlevels(txdb_gmx)[1:5]
}
```

# Rice (*Oryza sativa*) example

```{r load_osa}
osa_file <- file.path("plant_txdb_sqlite", "TxDb.Osativa.IRGSP.v62.sqlite")
if (!file.exists(osa_file)) {
  message("File not found. Run download chunk first.")
} else {
  txdb_osa <- loadDb(osa_file)
  head(genes(txdb_osa))
  seqlevels(txdb_osa)[1:5]
}
```

# Session information

```{r sessionInfo}
sessionInfo()
```
